---
title: "Extending roxygen2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending roxygen2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
library(roxygen2)
```

## Basics

Roxygen is extensible with user-defined roclets. 
It means that you can take advantage of Roxygen's parser in various ways. 
It is not a general tool for static analysis as only so-called 'blocks' are parsed, i.e. 
comment blocks that has the `#'` line prefix. 
On the other hand, the parser carries along information on the corresponding code 
which then can be used.
Thus, as we will demonstrate first, you can exploit the parser and use the information in blocks 
and corresponding code. 

Another option is to introduce custom tags similar to e.g. `@title`, `@param` etc. tags. 
We will cover this in the end of this vignette.

It is worth noting that it is not possible for a user to extend 
included roclets (e.g. the `rd` roclet) to allow for new tags. 
It is necessary to write a new roclet that can be used together with or instead of e.g. 
the included `rd` roclet.

## Roclets

Running a roclet consists of multiple parts as documented in `?roclet`. 
Very briefly, a roclet can implement a number of functions (please refer to `?roclet` for details):

* `roclet_tags()`: named list of tags whose values are tag parsing functions (optional to implement)
* `roclet_preprocess()`: called after blocks have been parsed but before code has been evaluated (optional to implement)
* `roclet_process()`: called after `@eval` tag has been processed (main functionality)
* `roclet_output()`: produces files on disk based on the result from `roclet_process()` (output functionality)

Depending on the roclet, there may be some work in managing files, ensuring that no existing files are over-written, deleted etc. That must be handled, too. 
Be sure not to change or delete something that you did not create with the explicit 
disclaimer that the content can be modified or deleted. 
Roxy does this by the header (first line of the file) containing "`Generated by roxygen2: do not edit by hand`".

A roclet will use a list of tag parsers to process the written documentation. 
Internally, roxygen creates a `roxy_tag()` which is then passed on to the 
tag parser specified.

### Custom roclets

This example will list the functions with inconsistent parameter documentation, e.g. missing `@param`s or too many `@param`s. 
This is caught when `R` is checking a package, but sometimes it is nice to catch is before that. 
The check will be a lot more simplistic than the one from `R` check, but the idea is to illustrate Roxygen's functionality.

Other examples of such functionality is to display 
functions with most lines (to avoid too long functions), 
functions with most parameters, 
inconsistent parameter naming and so on.

Let us focus on a block like this:

```{r}
txt1 <- "
#' Summing two numbers
#'
#' @param x A number
#' @param y Another number
f <- function(x, y) {
  x + y
}"
```

The output from the `rd` roclet will be:

```{r}
roc_proc_text(rd_roclet(), txt1)
```

Say that we forgot to document the `y` parameter:

```{r}
txt2 <- "
#' Summing two numbers
#'
#' @param x A number
f <- function(x, y) {
  x + y
}"
```

The output from the `rd` roclet will be:

```{r}
roc_proc_text(rd_roclet(), txt2)
```

That is in principle fine. But we do not discover that `y` is missing a `@param` (until checking the package). 
Thus, we make a roclet to help us catch this earlier. 
The input argument `blocks` to `roclet_process()` is a list of blocks. 
Below, we show the block from `txt2` to give you an idea of its structure:

```
List of 2
 $ title: chr "Summing two numbers"
 $ param:List of 2
  ..$ name       : chr "x"
  ..$ description: chr "A number"
 - attr(*, "filename")= chr "/tmp/RtmpZx52lH/file2498288d1e88"
 - attr(*, "location")= int [1:8] 5 1 7 1 1 1 5 7
 - attr(*, "call")= language f <- function(x, y) {     x + y ...
 - attr(*, "class")= chr "roxy_block"
 - attr(*, "object")=List of 3
  ..$ alias  : chr "f"
  ..$ value  :function (x, y)  
  ..$ methods: NULL
  ..- attr(*, "class")= chr [1:2] "function" "object"
```

Now, we can implement the roclet:

```{r}
param_roclet <- function() {
  roxygen2::roclet("param")
}

roclet_process.roclet_param <- function(x, blocks, env, base_path, global_options = list()) {
  warns <- list()
  
  for (block in blocks) {
    block_obj <- attr(block, "object")
    
    if (!is(block_obj, "function")) {
      next
    }
    
    fun_args <- formalArgs(block_obj$value)
    block_params <- block[names(block) == "param"]
    block_params_names <- sort(unname(sapply(block_params, function(x) x$name), force = TRUE))
  
    if (!isTRUE(all.equal(fun_args, block_params_names))) {
      func_name <- block_obj$alias
      
      missing_params <- setdiff(fun_args, block_params_names)
      toomany_params <- setdiff(block_params_names, fun_args)
      msg <- ""
      if (length(missing_params) > 0) {
        msg <- paste0(msg, "\n    - Missing @param's: ", paste0(missing_params, collapse = ", "))
      }
      if (length(toomany_params) > 0) {
        msg <- paste0(msg, "\n    + Too many @param's: ", paste0(toomany_params, collapse = ", "))
      }

      warn <- paste0("Function '", func_name, "' with title '", block$title, "': ", msg)
      warns <- c(warns, warn)
    }
  }
  
  return(warns)
}

roc_proc_text(param_roclet(), txt2)
```

Notice here that the roclet is only processed. 
The output is not made. 
To specify the output functionality you must implement a `roclet_output.roclet_param()` method.
It would probably go something like:

```{r}
roclet_output.roclet_param <- function(x, results, base_path, ...) {
  if (length(results) == 0L) {
    return(invisible(NULL))
  }
  
  cat("Functions with @param inconsistency:\n")
  cat(paste0("  * ", results, collapse = "\n"), sep = "")
  
  return(invisible(NULL))
}
```

We can now make complete examples of three different cases:

```{r}
# All good:
roclet_output(param_roclet(), roc_proc_text(param_roclet(), "
#' Summing two numbers
#'
#' @param x A number
#' @param y Another number
f <- function(x, y) {
  x + y
}"))
# Missing documentation of 'y' parameter
roclet_output(param_roclet(), roc_proc_text(param_roclet(), "
#' Summing two numbers
#'
#' @param x A number
f <- function(x, y) {
  x + y
}"))
# Additional documentation for 'z' parameter
roclet_output(param_roclet(), roc_proc_text(param_roclet(), "
#' Summing two numbers
#'
#' @param x A number
#' @param y Another number
#' @param z A third parameter
f <- function(x, y) {
  x + y
}"))
```


### Custom tags

In this example we will make a new `@memo` tag to enable printing the memos at the console when the roclet runs.
As you will see, testing a complete workflow of a roclet can be a bit cumbersome. 

We choose that the `@memo` has this syntax:

```
@memo [Headline] Description
```

As an example:

```
@memo [EFFICIENCY] Currently brute-force; find better algorithm.
```

The idea is that all memos with the same headline are grouped together.

```{r}
memo_roclet <- function() {
  roxygen2::roclet("memo")
}

# Based on roxygen2:::tag_name_description
tag_memo <- function(x) {
  #print(x)
  #print(str(x))
  
  # Check syntax
  stopifnot(grepl("^\\[.*\\].*$", x$val))
  
  memo_parsed <- stringi::stri_match(str = x$val, regex = "\\[(.*)\\](.*)")
  
  header <- memo_parsed[1L, 2L]
  msg <- memo_parsed[1L, 3L]
  
  x$val <- list(header = header, msg = msg)
  
  return(x)
}

roclet_tags.roclet_memo <- function(x) {
  list(memo = tag_memo)
}

roclet_process.roclet_memo <- function(x, blocks, env, base_path, global_options = list()) {
  memos <- list()
  
  print(str(blocks))
  
  for (block in blocks) {
    block_memos <- block[names(block) == "memo"]
    print(str(block_memos))
    
    if (length(block_memos) == 0L) {
      next
    }
    
    for (m in block_memos) {
      extra <- ""
      
      if (!is.null(attr(block, "object")) && 
          !is.null(attr(block, "object")$alias) &&
          !is.null(attr(block, "filename"))) {
        extra <- paste0(" @ '", attr(block, "object")$alias, "' [", attr(block, "filename"), "]")
      }
      
      msg <- paste0(m$msg, extra)
      memos[[m$header]] <- c(memos[[m$header]], msg)
    }
    
  }
  
  return(memos)
}

roclet_output.roclet_memo <- function(x, results, base_path, ...) {
  if (length(results) == 0L) {
    return(invisible(NULL))
  }
  
  cat("Memos:\n")
  
  for (i in seq_along(results)) {
    cat("  ", names(results)[i], ":\n", sep = "")
    cat(paste0("    - ", results[[i]], collapse = "\n"), sep = "")
    cat("\n")
  }
  
  return(invisible(NULL))
}

results <- roc_proc_text(memo_roclet(), "
#' @memo [TBI] Remember to implement this!
#' @memo [API] Check best API
f <- function(x, y) {
  # ...
}

#' @memo [API] Consider passing z option
g <- function(x, y) {
  # ...
}
", registry = roclet_tags.roclet_memo())
roclet_output(memo_roclet(), results)
```


