---
title: "Extending roxygen2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Extending roxygen2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

roxygen2 is extensible, and this vignette will show you how.
It starts with an introduction to the basic workflow of `roxygenize()` and the key data structures that power it. Then we'll show you how you can use its two extension points:

-   Add a new tag to generate a new top-level section in an `.Rd` file. This allows you to repeat yourself less when documenting your package. (See `vignette("reuse")` for other techniques.)

-   Add a new roclet. This lets you take full advantage of the computational machinery behind `roxygenize()` to compute anything you want or produce any artefact you can imagine.

```{r setup}
library(roxygen2)
```

## How `roxygenize()` works

You've probably used `roxygenize()` (or `devtools::document()`) a bunch without ever really thinking about what's going on behind the scenes.
But if you're going to extend roxygen2, you'll need to know exactly what's happening:

- Loads the package under roxygenizing, as well as any further packages ("packages" in `load_options()`).

- Parses all R files in the package, using available tags. 

- Finds the roclets (see `roclet()`) from its `roclets` argument or the "roclets" option (`load_options()`). It defaults to using the Collate "roclet", the Rd roclet, and NAMESPACE roclet, but you can also add your own.

- Runs the different methods of all those roclets, in order and independently: clean (`roclet_clean`), preprocess (`roclet_preprocess()`), process (`roclet_process()`), and output (`roclet_output()`). Only process and output are routinely used. For example, if you think of the Rd roclet, its process method digests information from the tag parsing, combines inherits, etc. to create the content of each documentation topic, and its output method writes those topics to disk.

## Key data structures

Before we dive into extending roxygen2, we need to first discuss two important data structures that power roxygen: tags and blocks.

### Tags

A tag (a list with S3 class `roxy_tag`) represents a single tag.
It has the following fields:

-   `tag`: the name of the tag.

-   `raw`: the raw contents of the tag (i.e. everything from the end of this tag to the beginning of the next).

-   `val`: the parsed value, which we'll come back to shortly.

-   `file` and `line`: the location of the tag in the package.
    Used with `roxy_tag_warning()` to produce informative error messages.

You *can* construct tag objects by hand with `roxy_tag()`:

```{r}
roxy_tag("name", "Hadley")
str(roxy_tag("name", "Hadley"))
```

However, you should rarely need to do so (except in tests), because you'll typically have them given to you in a block object, as you'll see shortly.

### Blocks

A block (a list with S3 class `roxy_block`) represents a single roxygen block.
It has the following fields:

-   `tags`: a list of `roxy_tags`.
-   `call`: the R code associated with the block (usually a function call).
-   `file` and `line`: the location of the R code.
-   `object`: the evaluated R object associated with the code.

The easiest way to see the basic structure of a `roxy_block()` is to generate one by parsing a roxygen block with `parse_text()`:

```{r}
text <- "
  #' This is a title
  #'
  #' This is the description.
  #'
  #' @param x,y A number
  #' @export
  f <- function(x, y) x + y
"

# parse_text() returns a list of blocks, so I extract the first
block <- parse_text(text)[[1]]
block
```

You'll notice that some of the tags didn't exist in the original block:

* `@title` and `@description` are extracted from the text that appears before the first explicit tag.
* `@usage` is generated automatically from the function formals.
* `@.formals` is an "internal" tag that doesn't generate any output but is used to pass some important data around.
* `@backref` stores the source location of the block so we can later record which `.R` files contributed to each `.Rd` file.

## Adding a new `.Rd` tag

The most common way to extend roxygen2 is to create a new tag that adds output to `.Rd` files.
This requires defining a few methods:

1.  Define a `roxy_tag_parse()` method that describes how to parse our new tag.

2.  Define a `roxy_tag_rd()` method that describes how to convert the tag into `.Rd` commands.

3.  If the tag's content is meant to appear in a custom section (as opposed to, say, the examples section), define a `format()` method that describes how to create the `.Rd` string.

To illustrate the basic idea, we'll create a new `@tip` tag that will create a bulleted list of tips about how to use a function.
The idea is to take something like this:

```{r}
#' @tip The mean of a logical vector is the proportion of `TRUE` values.
#' @tip You can compute means of dates and date-times!
```

That generates Rd like this:

``` latex
\section{Tips and tricks}{
\itemize{
  \item The mean of a logical vector is the proportion of \code{TRUE} values.
  \item You can compute means of dates and date-times!
}
}
```

The first step is to define a method for `roxy_tag_parse()` that describes how to parse the tag text.
The name of the class will be `roxy_tag_{tag}`, which in this case is `roxy_tag_tip`.
This function takes a `roxy_tag` as input, and its job is to set `x$val` to a convenient parsed value that will be used later by the roclet.
Here we want to process the text using Markdown so we can just use `tag_markdown()`:

```{r}
roxy_tag_parse.roxy_tag_tip <- function(x) {
  tag_markdown(x)
}
```

(There are lots of other built in options that you can read about in `?tag_markdown`.)

```{r, include = FALSE}
# Needed for vignette
registerS3method("roxy_tag_parse", "roxy_tag_tip", roxy_tag_parse.roxy_tag_tip)
```

We can check this works by using `parse_text()`:

```{r}
text <- "
  #' Title
  #'
  #' @tip The mean of a logical vector is the proportion of `TRUE` values.
  #' @tip You can compute means of dates and date-times!
  #' @md
  f <- function(x, y) {
    # ...
  }
"
block <- parse_text(text)[[1]]
block

str(block$tags[[2]])
```

Here I explicitly turn Markdown parsing on using `@md`; it's usually turned on for a package using roxygen2's `load_options()`.

Next, we define a method for `roxy_tag_rd()`, which must create an `rd_section()`.
We're going to create a new custom section called `tip`.
It will contain a character vector of tips:

```{r}
roxy_tag_rd.roxy_tag_tip <- function(x, base_path, env) {
  rd_section("tip", x$val)
}
```

```{r, include = FALSE}
# Needed for vignette
registerS3method("roxy_tag_rd", "roxy_tag_tip", roxy_tag_rd.roxy_tag_tip)
```

This additional layer is needed because there can be multiple tags of the same type in a single block, and multiple blocks can contribute to the same `.Rd` file.
The job of `rd_section()` is to combine all the tags into a single top-level Rd section.
Each tag generates an `rd_section` which is then combined with any previous section using `merge()`.
The default `merge.rd_section()` just concatenates the values together (`rd_section(x$type, c(x$value, y$value))`); you can override this method if you need more sophisticated behaviour.

We called the custom section "tip" just like our tag, but we needn't have done so: it really depends on how you want to map the input tags to output Rd code.

We then need to define a `format()` method to convert this object into text for the `.Rd` file:

```{r}
format.rd_section_tip <- function(x, ...) {
  paste0(
    "\\section{Tips and tricks}{\n",
    "\\itemize{\n",
    paste0("  \\item ", x$value, "\n", collapse = ""),
    "}\n",
    "}\n"
  )
}
```

```{r, include = FALSE}
# Needed for vignette
registerS3method("format", "rd_section_tip", format.rd_section_tip)
```

We can now try this out with `roclet_text()`:

```{r}
topic <- roc_proc_text(rd_roclet(), text)[[1]]
topic$get_section("tip")
```

Note that there is no namespacing so if you're defining multiple new tags I recommend using your package name as the common prefix.

### Using your new tag

Now that the three methods are created, we still need to make them available to `roxygenize()`. First, you need to export the method:

* If the package defining the tag _imports_ roxygen2, use `@export`.
* If the package defining the tag only _suggests_ roxygen2, use `@exportMethod`.

Next, you'll need to load the tag-defining package in the package where you want to use it. For example, if you created some new tags in {packageFoo}, and would like to use these tags in your documentation for {packageBar}, append this line to the `DESCRIPTION` of {packageBar}:

```
Config/roxygen2/packages: packageFoo
```

See `load_options()` for more details.

## Creating a new roclet

Creating a new roclet is usually a two part process.
First, you define new tags that your roclet will work with, unless your roclet only needs information from existing tags, or only needs the path to the package source[^vignette].
Second, you define a roclet that tells `roxygenize()` what to compute and produce based on this information.

[^vignette]: For example, the no-longer recommended `vignette_roclet()` only needs the path to the package source as input; it does not use information from the tag parsing step. Or the {roxylint} package only uses existing tags; its job is to warn you about suboptimal roxygen2 style.

### Custom tags

In this example we will make a new `@memo` tag which helps you to remember what you're planning to work on in the future by displaying notes when you document your package.
We choose this syntax for `@memo`:

```{r}
#' @memo [Headline] Description
```

For example:

```{r}
#' @memo [EFFICIENCY] Currently brute-force; find better algorithm.
```

As above, we first define a parse method. This time we use custom format based on a regular expression:

```{r}
roxy_tag_parse.roxy_tag_memo <- function(x) {
  if (!grepl("^\\[.*\\].*$", x$raw)) {
    roxy_tag_warning(x, "Invalid memo format")
    return()
  }

  parsed <- stringi::stri_match(str = x$raw, regex = "\\[(.*)\\](.*)")[1, ]

  x$val <- list(
    header = parsed[[2]],
    message = parsed[[3]]
  )
  x
}
```

```{r, include = FALSE}
# Needed for vignette
registerS3method(
  "roxy_tag_parse",
  "roxy_tag_memo",
  roxy_tag_parse.roxy_tag_memo
)
```

Then we check it works with `parse_text()`:

```{r}
text <- "
  #' @memo [TBI] Remember to implement this!
  #' @memo [API] Check best API
  f <- function(x, y) {
    # ...
  }
"
block <- parse_text(text)[[1]]
block

str(block$tags[[1]])
```

We don't need a format method because our tag won't be used to produce Rd sections[^tag].

[^tag]: Some tags in roxygen2 itself, like `@importFrom`, are not meant for the [`rd_roclet()`] (`@importFrom` is meant for the `namespace_roclet()`).

### The roclet

Next, we create a constructor for the roclet, which uses `roclet()`.
Our `memo` roclet doesn't have any options so this is very simple:

```{r}
memo_roclet <- function() {
  roclet("memo")
}
```

To give the roclet behaviour, you need to define methods.
There are two methods that almost every roclet will use:

-   `roclet_process()` is called with a list of blocks, and returns an object of your choosing.

-   `roclet_output()` produces side-effects (usually writing to disk) using the result from `roclet_process()`.

For this roclet, we'll have `roclet_process()` collect all the memo tags into a named list:

```{r}
roclet_process.roclet_memo <- function(x, blocks, env, base_path) {
  results <- list()

  for (block in blocks) {
    tags <- block_get_tags(block, "memo")

    for (tag in tags) {
      msg <- paste0("[", tag$file, ":", tag$line, "] ", tag$val$message)
      results[[tag$val$header]] <- c(results[[tag$val$header]], msg)
    }
  }

  results
}
```

And then have `roclet_output()` print them to the screen:

```{r}
roclet_output.roclet_memo <- function(x, results, base_path, ...) {
  for (header in names(results)) {
    messages <- results[[header]]
    cat(paste0(header, ": ", "\n"))
    cat(paste0(" * ", messages, "\n", collapse = ""))
  }

  invisible(NULL)
}
```

```{r, include = FALSE}
# Needed for vignette
registerS3method("roclet_process", "roclet_memo", roclet_process.roclet_memo)
registerS3method("roclet_output", "roclet_memo", roclet_output.roclet_memo)
```

Then you can test if it works by using `roc_proc_text()`:

```{r}
results <- roc_proc_text(
  memo_roclet(),
  "
#' @memo [TBI] Remember to implement this!
#' @memo [API] Check best API
f <- function(x, y) {
  # ...
}

#' @memo [API] Consider passing z option
g <- function(x, y) {
  # ...
}
"
)
roclet_output(memo_roclet(), results)
```

## Adding a roclet to your workflow

To use a roclet when developing a package, call

```r
roxygen2::roxygenize(roclets = "yourPackage::roclet")
```

where `yourPackage::roclet` is the function which creates the roclet, e.g. `memo_roclet` above.

You can also add the roclet to the target package's DESCRIPTION file, like this:

```r
Config/roxygen2/roclets: collate, rd, namespace, yourPackage::roclet
```

See `load_options()` for more details.

Your package only needs to be installed when the user documents the package, which doesn't correspond precisely to any field in the `DESCRIPTION`. However, you can think of it as a development dependency and hence put it in the `Suggests:` field:

```r
usethis::use_package("yourPackage", type = "Suggests")
```

You don't have to do this, but it will help other developers working on the target package.

## Conclusion: further extension ideas

This vignette is quite rough, so you might want to also read some roxygen2 source code to understand all the extension points.

Do not hesitate to also look for examples of roxygen2 extensions.
For instance, the [roxygenlabs](https://github.com/gaborcsardi/roxygenlabs) package (former incubator of roxygen2 features) used a third extension point: it _extended_ the Rd roclet with further methods, thus creating a supercharged Rd roclet rather than a brand-new roclet.
Or, the [plumber2](https://github.com/posit-dev/plumber2) package only uses the parsing features from roxygen2, and does not use `roxygenize()` at all.
