---
title: "Text formatting reference sheet"
author: "Hadley Wickham"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Text formatting reference sheet}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

Within roxygen tags, you use `.Rd` syntax to format text. This vignette shows you examples of the most important commands. The full details are described in [R extensions](https://cran.r-project.org/doc/manuals/R-exts.html#Marking-text).

Note that `\` and `%` are special characters. To insert literals, escape with a backslash: `\\`, `\%`.

## Character formatting

* `\emph{italics}`

* `\strong{bold}`

* `\code{r_function_call(with = "arguments")}`, `\code{NULL}`, `\code{TRUE}`

* `\pkg{package_name}`

## Links

To other documentation:

* `\code{\link{function}}`: function in this package

* `\code{\link[MASS]{stats}}`: function in another package

* `\link[=dest]{name}`: link to dest, but show name

* `\linkS4class{abc}`: link to an S4 class

To the web:

* `\url{http://rstudio.com}`

* `\href{http://rstudio.com}{Rstudio}`

* `\email{hadley@@rstudio.com}` (note the doubled `@`)

## Lists

* Ordered (numbered) lists:

    ```{r}
    #' \enumerate{
    #'   \item First item
    #'   \item Second item
    #' }
    ```

* Unordered (bulleted) lists

    ```{r}
    #' \itemize{
    #'   \item First item
    #'   \item Second item
    #' }
    ```

* Definition (named) lists

    ```{r}
    #' \describe{
    #'   \item{One}{First item}
    #'   \item{Two}{Second item}
    #' }
    ```

## Mathematics

Standard LaTeX (with no extensions):

* `\eqn{a + b}`: inline eqution

* `\deqn{a + b}`: display (block) equation

## Tables

Tables are created with `\tabular{}`. It has two arguments:

1. Column alignment, specified by letter for each column (`l` = left, `r` = right,
   `c` = centre.)

2. Table contents, with columns separated by `\tab` and rows by `\cr`.

The following function turns an R data frame into into the correct format. It ignores column and row names, but should get you started.

```{r}
tabular <- function(df, ...) {
  stopifnot(is.data.frame(df))

  align <- function(x) if (is.numeric(x)) "r" else "l"
  col_align <- vapply(df, align, character(1))

  cols <- lapply(df, format, ...)
  contents <- do.call("paste",
    c(cols, list(sep = " \\tab ", collapse = "\\cr\n  ")))

  paste("\\tabular{", paste(col_align, collapse = ""), "}{\n  ",
    contents, "\n}\n", sep = "")
}

cat(tabular(mtcars[1:5, 1:5]))
```
