---
title: "Managing imports and exports"
description: >
  Generating the `NAMESPACE` file with roxygen2.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing imports and exports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

The package `NAMESPACE` is one of the most confusing parts of building a package.
roxygen2 aims to make it as easy as possible to build a package that is a well-behaved member of the R ecosystem.
This is a little frustrating at first, but soon becomes second-nature.

## Exports

In order for your users to use a function[^1] in your package, you must **export** it.
In most cases, you can just use the `@export` tag, and roxygen2 will automatically figure out which `NAMESPACE` directive (e.g. `export()`, `exportS3method()`, `exportClasses()`, and `exportMethods()`) you need.

[^1]: Including S3 and S4 generics and methods.

### Functions

Functions should be exported if you want your users to available.
If you export a function, you must also document it, and since other people will use it, you need to be careful if you later change the function interface.

```{r}
#' Add two numbers together
#' 
#' @param x,y Two numbers
#' @export
add <- function(x, y) {
  x + y
}
```

### S3

S3 generics work like regular R functions so export them using the advice above: if you want users to call this function, export; if it's purely for internal use, don't export it.

```{r}
#' Take an object to bizarro world
#' 
#' @param x A vector.
#' @export
bizarro <- function(x, ...) {
  UseMethod("bizarro")
}
```

While S3 methods are regular functions with a special naming scheme, their "export" works a bit differently.
S3 methods are exported only in the sense that when you call the generic with the appropriate method; a user can't directly access the function definition by typing the name of the generic.
A more technically correctly term would be to say that the method is **registered** so that the generics can find it.
You must register, i.e. `@export`, every S3 method regardless of whether or not the generic is exported, and roxygen2 will warn you if you have forgotten.

```{r}
#' @export
bizarro.character <- function(x, ...) {
  letters <- strsplit(x, "")
  letters_rev <- lapply(letters, rev)
  vapply(letters_rev, paste, collapse = "", FUN.VALUE = character(1))
}
```

You can document S3 methods, but you're not required to.
If you document a method, you have three choices.
If the method is particularly complex or has many arguments that the generic does not, you can document it in its own file.
Otherwise you can use `@rdname` to document it with either the generic or the class.

```{r}
#' Take an object to bizarro world
#' 
#' @description
#' This is an S3 generic. This package provides methods for the 
#' following classes:
#' 
#' * `character`: reverses the order of the letters in each element of 
#'    the vector.
#' 
#' @param x A vector.
#' @export
bizarro <- function(x, ...) {
  UseMethod("bizarro")
}

#' @export
#' @rdname bizarro
bizarro.character <- function(x, ...) {
  letters <- strsplit(x, "")
  letters_rev <- lapply(letters, rev)
  vapply(letters_rev, paste, collapse = "", FUN.VALUE = character(1))
}
```

Typically, you will write methods for generics that are either defined in the current package or a package that is a hard dependency[^2] of your package.
Sometimes, however, you will want to write a method for a suggested dependency.
In this case, `@export` will not work because it relies on being able to see the generic.
Instead, use `@exportS3Method`. This will use "delayed" method registration, which means the method will only be registered when the suggested package is loaded.

[^2]: i.e. listed the `Imports` or `Depends` fields in your `DESCRIPTION`.

```{r}
#' @exportS3Method pkg::generic
generic.foo <- function(x, ...) {
}
```

### S4

-   **Classes**: export the class object if you want others to be able to extend it.

-   **Generics:** treat it like a function and `@export` if user facing.

-   **Methods**: you only need to `@export` a method, if the generic lives in another package.

### Datasets

Note that datasets should never be exported as they use a different mechanism.
Instead, datasets will either be automatically exported if you set `LazyData: true` in your `DESCRIPTION`, or made available after calling `data()` if not.

### Manual exports

If `@export` does not automatically generate the correct directive when, you can use one of the tags below to exercise greater control:

-   `@export foo` generates `export(foo)`
-   `@exportS3Method generic method` generates `S3method(generic, method)`
-   `@exportClass foo` generates `exportClasses(foo)`
-   `@exportMethod foo` generates `exportMethods(foo)`
-   `@exportPattern foo` generates `exportPattern(foo)`

For even more specialised cases you can use `@rawNamespace code` which inserts `code` literally into the `NAMESPACE`.
If you need to automate this, `@evalNamespace fun()` will evaluate `fun()` in the package environment and insert the results into `NAMESPACE`.
Because `evalNamespace()` is run in the package environment, it can only generate exports, not imports.

## Imports

The `NAMESPACE` also controls which functions from other packages are made available to your package.

### Functions

If you are using just a few functions from another package, we recommending adding the package to the `Imports:` field of the `DESCRIPTION` file and calling the functions explicitly using `::`, e.g., `pkg::fun()`.

``` r
my_function <- function(x, y) {
  pkg::fun(x) * y
}
```

If the repetition of the package name becomes annoying you can `@importFrom` and drop the `::`:

``` r
#' @importFrom pkg fun 
my_function <- function(x, y) {
  fun(x) * y
}
```

Imports affect every function in a package, so it's common to collect them in a central place, like `{packagename}-package.R`.
This is automated by `usethis::use_import_from()`.

```{r}
#' @importFrom pkg fun1 fun2
#' @importFrom pkg2 fun3
#' @importFrom pkg3 fun4
NULL
```

Note the use of `NULL` here: you must provide something for roxygen2 to document, so we use `NULL` as place holder.

It is possible, but not generally recommended to import all functions from a package with `@import package`.
This is risky if you import functions from more than one package, because while it might be ok today, in the future the packages might end up with a function having the same name, and your users will get a warning every time your package is loaded.

### S3

S3 generics are just functions, so the same rules for functions apply.
S3 methods always accompany the generic, so as long as you can access the generic (either implicitly or explicitly), the methods will also be available.
In other words, you don't need to do anything special for S3 methods.
As long as you've imported the generic, all the methods will also be available.

### S4

-   To use classes defined in another package, place `@importClassesFrom package ClassA ClassB ...` next to the classes that inherit from the imported classes, or next to the methods that implement a generic for the imported classes.
-   To use generics defined in another package, place `@importMethodsFrom package GenericA GenericB ...` next to the methods that use the imported generics.

### Compiled code

To import compiled code from another package, use `@useDynLib`

-   `@useDynLib package` imports all compiled functions.

-   `@useDynLib package routinea routineb` imports selected compiled functions.

-   Any `@useDynLib` specification containing a comma, e.g. `@useDynLib mypackage, .registration = TRUE` will be inserted as is into the the NAMESPACE, e.g. `useDynLib(mypackage, .registration = TRUE)`
