---
title: "Managing imports and exports"
description: >
  Generating the `NAMESPACE` file with roxygen2.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Managing imports and exports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

The package `NAMESPACE` is one of the most confusing parts of building a package.
roxygen2 aims to make it as easy as possible to build a package that is a well-behaved member of the R ecosystem.
This is a little frustrating at first, but soon becomes second-nature.

## Exports

In order for your users to use a function[^1] from your package, you must **export** it.
In most cases, you can just use the `@export` tag, and roxygen2 will automatically figure out which `NAMESPACE` directive (i.e. `export()`, `S3method()`, `exportClasses()`, or `exportMethods()`) you need.

[^1]: Including S3/S4/S7 generics and constructors.

For details and examples of exporting functions, and S3, S4, and S7 classes/generics/methods, see the corresponding vignettes:

- `vignette("rd-functions")` for regular functions.
- `vignette("rd-S3")` for S3 generics and methods, including delayed registration with `@exportS3Method`.
- `vignette("rd-S4")` for S4 generics, classes, and methods.
- S7 methods are registered at load time, not through `NAMESPACE`; see `vignette("rd-S7")`.

Datasets should never be exported as they are not found in `NAMESPACE`.
Learn more in `vignette("rd-datasets")`.

### Manual exports

If `@export` does not automatically generate the correct `NAMESPACE` directive, you can use one of the tags below to exercise greater control:

- `@export foo` generates `export(foo)`
- `@exportS3Method generic method` generates `S3method(generic, method)`
- `@exportClass foo` generates `exportClasses(foo)`
- `@exportMethod foo` generates `exportMethods(foo)`
- `@exportPattern foo` generates `exportPattern(foo)`

For even more specialised cases you can use `@rawNamespace code` which inserts `code` literally into the `NAMESPACE`.
This is useful if you need a conditional import or export, e.g.

```{r}
# From dplyr:
#' @rawNamespace import(vctrs, except = data_frame)
NULL

# From backports:
#' @rawNamespace if (getRversion() < "4.0.0") export(stopifnot)
NULL
```

If you need to automate this, `@evalNamespace fun()` will evaluate `fun()` in your package environment and insert the results into `NAMESPACE`.
Note that because `evalNamespace()` is run in the package environment, it can only generate exports, not imports.

## Imports

The `NAMESPACE` also controls which functions from other packages are made available to your package.

### Functions

If you are using just a few functions from another package, we recommending adding the package to the `Imports:` field of the `DESCRIPTION` file and calling the functions explicitly using `::`, e.g., `pkg::fun()`.

```r
my_function <- function(x, y) {
  pkg::fun(x) * y
}
```

If the repetition of the package name becomes annoying you can `@importFrom` and drop the `::`:

```r
#' @importFrom pkg fun
my_function <- function(x, y) {
  fun(x) * y
}
```

Imports affect every function in a package, so it's common to collect them in a central place, like `{packagename}-package.R`.
This is automated by `usethis::use_import_from()`.

```{r}
#' @importFrom pkg fun1 fun2
#' @importFrom pkg2 fun3
#' @importFrom pkg3 fun4
NULL
```

Note the use of `NULL` here: you must provide something for roxygen2 to document, so we use `NULL` as placeholder.

It is possible, but not generally recommended to import all functions from a package with `@import package`.
This is risky if you import functions from more than one package, because while it might be ok today, in the future the packages might end up with a function having the same name, and your users will get a warning every time your package is loaded.

### S3

S3 generics are just functions, so the same rules for functions apply.
S3 methods always accompany the generic, so as long as you can access the generic, the methods will also be available.
In other words, you don't need to do anything to import an S3 method.

### S4

- To use classes defined in another package, place `@importClassesFrom package ClassA ClassB ...` next to the classes that inherit from the imported classes, or next to the methods that implement a generic for the imported classes.
- To use generics defined in another package, place `@importMethodsFrom package GenericA GenericB ...` next to the methods that use the imported generics.

### Compiled code

To import compiled code from another package, use `@useDynLib`

- `@useDynLib package` imports all compiled functions.

- `@useDynLib package routinea routineb` imports selected compiled functions.

- Any `@useDynLib` specification containing a comma, e.g. `@useDynLib mypackage, .registration = TRUE` will be inserted as is into the the NAMESPACE, e.g. `useDynLib(mypackage, .registration = TRUE)`
