---
title: "Documenting datasets"
description: >
  How to document datasets stored in `data/`.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting datasets}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Datasets are stored in `data/`, not as regular R objects in the package.
This means you need to document them in a slightly different way: instead of documenting the data directly, you quote the dataset's name.
For example, this is the roxygen2 block used for `ggplot2::diamonds`:

```{r}
#| eval: false
#' Prices of over 50,000 round cut diamonds
#'
#' A dataset containing the prices and other attributes of almost 54,000
#'  diamonds. The variables are as follows:
#'
#' @format A data frame with 53940 rows and 10 variables:
#' \describe{
#'   \item{price}{price in US dollars ($326--$18,823)}
#'   \item{carat}{weight of the diamond (0.2--5.01)}
#'   \item{cut}{quality of the cut (Fair, Good, Very Good, Premium, Ideal)}
#'   \item{color}{diamond colour, from D (best) to J (worst)}
#'   \item{clarity}{a measurement of how clear the diamond is (I1 (worst), SI2,
#'     SI1, VS2, VS1, VVS2, VVS1, IF (best))}
#'   \item{x}{length in mm (0--10.74)}
#'   \item{y}{width in mm (0--58.9)}
#'   \item{z}{depth in mm (0--31.8)}
#'   \item{depth}{total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43--79)}
#'   \item{table}{width of top of diamond relative to widest point (43--95)}
#' }
#'
#' @source {ggplot2} tidyverse R package.
"diamonds"
```

Datasets should never be exported with `@export` because they are not found in the `NAMESPACE`.
Instead, datasets will either be automatically available if you set `LazyData: true` in your `DESCRIPTION`, or available after calling `data()` if not.
This field also affects the default usage.
If you have `LazyData: true`, the usage will be just the dataset name (e.g. `diamonds`).
Otherwise, the usage will be wrapped in `data()` (e.g. `data(diamonds)`).

Note the use of two additional tags that are particularly useful for documenting data:

- `@format`, which gives an overview of the structure of the dataset.
  This should include a **definition list** that describes each variable.
  There's currently no way to generate this with Markdown, so this is one of the few places you'll need to Rd markup directly.

- `@source` where you got the data form, often a URL.
