---
title: "Documenting functions"
description: >
  The basics of roxygen2 tags and how to use them for
  documenting functions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

This vignette describes the most important roxygen2 tags for documenting functions.
See `vignette("reuse")` for reusing documentation across topics.

## Exporting

For a user to use a function in your package, you must **export** it with `@export`.
If you export a function, you must also document it, and since other people will use it, you need to be careful if you later change the function interface.
We usually put `@export` in between `@returns` and `@examples`:

```{r}
#| eval: false
#' Add two numbers together
#'
#' @param x,y A pair of numbers.
#' @returns A number.
#' @export
#' @examples
#' 1 + 2
add <- function(x, y) {
  x + y
}
```

You'll learn what all these are pieces are next!

## The introduction

Each documentation block starts with some text which defines the title, the description, and the details.
Here's an example showing what the documentation for `sum()` might look like if it had been written with roxygen:

```{r}
#' Sum of vector elements
#'
#' `sum` returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the [Summary()] group generic. For this to work properly,
#' the arguments `...` should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

This introductory block is broken up as follows:

- The first sentence is the **title**: that's what you see when you look at `help(package = mypackage)` and is shown at the top of each help file.
  It should generally fit on one line, be written in sentence case, and not end in a full stop.

- The second paragraph is the **description**: this comes first in the documentation and should briefly describe what the function does.

- The third and subsequent paragraphs go into the **details**: this is a (often long) section that comes after the argument description and should provide any other important details of how the function operates.
  The details are optional.

### Title

The title appears in package indexes and search results, so think about how users will find your function.
Use synonyms to describe what the function does.
For example, dplyr uses titles like:

- "Keep rows that match a condition" (`filter()`)
- "Create, modify, and delete columns" (`mutate()`)
- "Summarise each group down to one row" (`summarise()`)

### Description

The description should briefly summarize the goal of the function, usually in one paragraph.
This can be challenging for simple functions, because it can feel like you’re just repeating the title of the function.
Try to find a slightly different wording, if you can.
It’s okay if this feels a little repetitive; it’s often useful for users to see the same thing expressed in two different ways.
It’s a little extra work, but the extra effort is often worth it.

```{r}
#| eval: false
#' Detect the presence/absence of a match
#'
#' `str_detect()` returns a logical vector with `TRUE` for each element of
#' `string` that matches `pattern` and `FALSE` otherwise. It's equivalent to
#' `grepl(pattern, string)`.
```

If you need a multi-paragraph description, use an explicit `@description` tag:

```{r}
#| eval: false
#' View strings and matches
#'
#' @description
#' `str_view()` is used to print the underlying representation of a string and
#' to see how a `pattern` matches.
#'
#' Matches are surrounded by `<>` and unusual whitespace (i.e. all whitespace
#' apart from `" "` and `"\n"`) are surrounded by `{}` and escaped.
```

Basically, if you’re going to include an empty line in your description, you’ll need to use an explicit `@description` tag.

### Details

The details section is optional and comes after the argument description in the rendered help.
Use informative markdown headings to break up long details.

```{r}
#' Sum of vector elements
#'
#' @description
#' `sum` returns the sum of all the values present in its arguments.
#'
#' @details
#' This is a generic function: methods can be defined for it directly
#' or via the [Summary()] group generic. For this to work properly,
#' the arguments `...` should be unnamed, and dispatch is on the
#' first argument.
```

## Required tags

Functions are the most commonly documented objects.
Functions require three tags: `@param`, `@returns`, and `@examples`.

### Inputs

Use `@param name description` to describe each input to the function.
The description should provide a succinct summary of the parameter type (e.g. a string, a numeric vector), and if not obvious from the name, what the parameter does.
The description is a sentence so should start with a capital letter and end with a full stop.
It can span multiple lines (or even paragraphs) if necessary.
All parameters must be documented.

```{r}
#| eval: false
#' @param pattern Pattern to look for.
#'
#'   The default interpretation is a regular expression, as described in
#'   `vignette("regular-expressions")`. Use [regex()] for finer control of the
#'   matching behaviour.
#'
#' @param string Input vector. Either a character vector, or something
#'   coercible to one.
```

It's a good idea to describe the default value in the `@param` description, particularly when the default is non-obvious or the signature and the rendered argument documentation are far apart.

```{r}
#| eval: false
#' @param na.rm Remove missing values? If `FALSE` (the default), the result
#'   will be `NA` if any element of `string` is `NA`.
```

For arguments with a small fixed set of possible values, list them in the description:

```{r}
#| eval: false
#' @param side Side on which to remove whitespace: `"left"`, `"right"`, or
#'   `"both"` (the default).
```

If each value needs more explanation, use a bulleted list:

```{r}
#| eval: false
#' @param whitespace_only A boolean.
#'   * `TRUE` (the default): wrapping will only occur at whitespace.
#'   * `FALSE`: can break on any non-word character (e.g. `/`, `-`).
```

If two or more arguments are tightly coupled, you can document them in one place by separating the names with commas (no spaces).
For example, to document both `x` and `y`, you can say `@param x,y Numeric vectors`.

### Outputs

`@returns description` describes the output from the function.
Briefly describe the type/shape of the output, not the details.

All functions must have a documented return value for initial CRAN submission.

For simple cases, a single sentence suffices:

```{r}
#| eval: false
#' @returns A logical vector the same length as `string`.
```

For functions returning data frames, describe how the output relates to the input:

```{r}
#| eval: false
#' @returns
#' An object of the same type as `.data`. The output has the following
#' properties:
#'
#' * Rows are a subset of the input, but appear in the same order.
#' * Columns are not modified.
#' * The number of groups may be reduced (if `.preserve` is not `TRUE`).
#' * Data frame attributes are preserved.
```

### Examples

`@examples` provides executable R code showing how to use the function in practice.
This is a very important part of the documentation because many people look at the examples before reading anything.
Example code must work without errors as it is run automatically as part of `R CMD check`.

Focus on demonstrating the most important features with realistic, typical usage.
Use comments to break examples into sections:

```{r}
#| eval: false
#' @examples
#' fruit <- c("apple", "banana", "pear", "pineapple")
#' str_detect(fruit, "a")
#' str_detect(fruit, "^a")
#' str_detect(fruit, "a$")
#'
#' # Also vectorised over pattern
#' str_detect("aecfg", letters)
```

Examples must be self-contained and not change the user's state: if you modify `options()`, reset them; create files in `tempdir()` and delete them; don't change the working directory or write to the clipboard.

For the purpose of illustration, it's often useful to include code that causes an error.
`try()` is the best way to do this, because users will see the error message:

```{r}
#| eval: false
#' @examples
#' # Row sizes must be compatible when column-binding
#' try(bind_cols(tibble(x = 1:3), tibble(y = 1:2)))
```

You can also use `\dontrun{}` to prevent code from executing, but `try()` is generally preferred.

For code that only works in certain environments, use `@examplesIf`:

```r
#' @examplesIf interactive()
#' browseURL("https://roxygen2.r-lib.org")
```

This generates

```
\examples{
\dontshow{if (interactive() (if (getRversion() >= "3.4") withAutoprint else force)(\{ # examplesIf}
gh_organizations(since = 42)
\dontshow{\}) # examplesIf}
}
```

This way, the code evaluating whether the example can be run is not shown to users reading the help, but it still prevents R CMD check failures.
`@examplesIf` is preferred over wrapping code in `if()` because users don't see the machinery, and unlike `\dontrun{}`, the code still runs when the condition is met.

Instead of including examples directly in the documentation, you can put them in separate files and use `@example path/relative/to/package/root` to insert them into the documentation.

All functions must have examples for initial CRAN submission.

### Usage

In most cases, the function usage (which appears beneath the description in the generated docs) will be automatically derived from the function specification.
For the cases where it is not, please [file an issue](https://github.com/r-lib/roxygen2/issues) and use `@usage` to override the default with what you want.
If you want to suppress the usage altogether (which is sometimes useful for internal or deprecated functions), you can use `@usage NULL`.
