---
title: "Documenting other objects"
description: > 
  How to document datasets, packages, and the classes,
  generics, and methods of S3, S4, and R6.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting other objects}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Datasets

Datasets are stored in `data/`, not as regular R objects in the package.
This means you need to document them in a slightly different way: instead of documenting the data directly, you quote the dataset's name. For example, this is the roxygen2 block used for `ggplot2::diamonds`:

```{r}
#| eval: false

#' Prices of over 50,000 round cut diamonds
#'
#' A dataset containing the prices and other attributes of almost 54,000
#'  diamonds. The variables are as follows:
#'
#' @format A data frame with 53940 rows and 10 variables:
#' \describe{
#'   \item{price}{price in US dollars ($326--$18,823)}
#'   \item{carat}{weight of the diamond (0.2--5.01)}
#'   \item{cut}{quality of the cut (Fair, Good, Very Good, Premium, Ideal)}
#'   \item{color}{diamond colour, from D (best) to J (worst)}
#'   \item{clarity}{a measurement of how clear the diamond is (I1 (worst), SI2,
#'     SI1, VS2, VS1, VVS2, VVS1, IF (best))}
#'   \item{x}{length in mm (0--10.74)}
#'   \item{y}{width in mm (0--58.9)}
#'   \item{z}{depth in mm (0--31.8)}
#'   \item{depth}{total depth percentage = z / mean(x, y) = 2 * z / (x + y) (43--79)}
#'   \item{table}{width of top of diamond relative to widest point (43--95)}
#' }
#' 
#' @source {ggplot2} tidyverse R package.
"diamonds"
```

The default usage depends on the `LazyData` setting in your `DESCRIPTION`. If you have `LazyData: true`, the usage will be just the dataset name (e.g. `diamonds`). Otherwise, the usage will be wrapped in `data()` (e.g. `data(diamonds)`).

Note the use of two additional tags that are particularly useful for documenting data:

-   `@format`, which gives an overview of the structure of the dataset.
    This should include a **definition list** that describes each variable.
    There's currently no way to generate this with Markdown, so this is one of the few places you'll need to Rd markup directly.

-   `@source` where you got the data form, often a URL.

## Packages

As well as documenting every object inside the package, you can also document the package itself by documenting the special sentinel `"_PACKAGE"`.
This automatically includes information parsed from the `DESCRIPTION`, including title, description, list of authors, and useful URLs.

We recommend placing package documentation in `{pkgname}-package.R`, and have `@keywords internal`.
Use `usethis::use_package_doc()` to set up automatically.

Here's an example:

```{r}
#| eval: false

#' @keywords internal 
"_PACKAGE"
```

Package documentation is a good place to put `# Package options` that documents options used by the package.

Some notes:

-   By default, aliases will be added so that both `?pkgname` and `package?pkgname` will find the package help.
    If there's an existing function called `pkgname`, use `@aliases {pkgname}-package NULL` to override the default.

-   Use `@references` to point to published material about the package that users might find helpful.

## S3

Documenting [S3 code](https://adv-r.hadley.nz/s3.html) well requires thinking about three things:

-   **Generics**: mention that the function is a generic and list the
    available methods.
-   **Methods**: link back to the generic; only document individually
    when the method has unique behavior or arguments.
-   **Classes**: document the constructor.

### Generics

S3 **generics** are regular functions, so document them as such.
The documentation should mention that the function is a generic, because this
tells the reader that the behavior may vary depending on the input and that
they can write their own methods. For simple generics, you can do this
in the description:

```{r}
#| eval: false

#' Frobnpolicate an object
#'
#' @description
#' `frobnpolicate()` is an S3 generic that ..., with methods available for
#' the following classes:
#' 
#' `r doclisting::methods_inline("frobnpolicate")`
```

For more complicated generics, you can use a `# Methods` section to
provide more detail:

```{r}
#| eval: false

#' Frobnpolicate an object
#'
#' @description
#' `frobnpolicate()` does ...
#'
#' # Methods
#' `frobnpolicate()` is an S3 generic with methods available for the following 
#' classes:
#' 
#' `r doclisting::methods_list("frobnpolicate")`
```

You might also want to include a section that provides additional details for developers implementing their own methods.

Both examples above use the [doclisting](https://doclisting.r-lib.org/)
package to automatically generate a list of methods, with links to their
help topics. These examples use inline R code (`` `r ` ``), which generates the list at documentation time (i.e. when you run
`devtools::document()`). This only requires including doclisting in `Suggests`.

If you want the list to dynamically reflect all methods that are currently registered
(including methods registered by other packages), use an Rd code block
(`` `Rd ` ``) instead:

```{r}
#| eval: false

#' `Rd doclisting::methods_list("frobnpolicate")`
```

Using `` `Rd ` `` requires doclisting in `Imports`, and you'll need a dummy call to eliminate the `R CMD check` NOTE about unused imports:

```{r}
#| eval: false

ignore_unused_imports <- function() {
  doclisting::methods_list
}
```


### Classes

S3 **classes** have no formal definition, so document the [constructor](https://adv-r.hadley.nz/s3.html#s3-constructor).
Note that you don't need to list methods for a class: in S3, methods
belong to generics, not to classes. Users should look at the generic's
help page to learn about available methods.

### Methods

It is your choice whether or not to document S3 **methods**.
Generally, it's not necessary to document straightforward methods for
common generics like `print()`. (You should, however, always `@export`
S3 methods, even internal ones.)

It's good practice, however, to document methods that have unique behavior or arguments. For example, the clock package documents the methods for
[`date_group()`](https://clock.r-lib.org/reference/date_group.html) on
their own pages ([Date](https://clock.r-lib.org/reference/date-group.html),
[POSIXt](https://clock.r-lib.org/reference/posixt-group.html)) because the
methods accept different `precision` values and have type-specific return
value semantics and examples. The generic page serves only as a signpost
linking to the individual method pages.

When documenting a method, always include a link back to the generic
using `[generic_name()]` so the reader can easily find the full
documentation and other methods.

### Method disambiguation

Generally, roxygen2 will automatically figure out the generic that the
method belongs to, and you should only need to use `@method` if there
is ambiguity. For example, is `all.equal.data.frame()` the
`equal.data.frame` method for `all()`, or the `data.frame` method for
`all.equal()`? If this happens to you, disambiguate with (e.g.)
`@method all.equal data.frame`.

## S4

S4 **generics** are also functions, so document them as such.

Document **S4 classes** by adding a roxygen block before `setClass()`.
Use `@slot` to document the slots of the class.
Here's a simple example:

```{r}
#| eval: false

#' An S4 class to represent a bank account
#'
#' @slot balance A length-one numeric vector
Account <- setClass("Account",
  slots = list(balance = "numeric")
)
```

S4 **methods** are a little more complicated.
Unlike S3 methods, all S4 methods must be documented.
You can document them in three places:

-   In the class.
    Most appropriate if the corresponding generic uses single dispatch and you created the class.

-   In the generic.
    Most appropriate if the generic uses multiple dispatches and you control it.

-   In its own file.
    Most appropriate if the method is complex.
    or the either two options don't apply.

Use either `@rdname` or `@describeIn` to control where method documentation goes.
See the next section for more details.

## R6

-   R6 methods can be documented in-line, i.e. the method's documentation comments come right before the definition of the method.

-   Method documentation can use the `@description`, `@details`, `@param`, `@return` and `@examples` tags.
    These are used to create a subsection for the method, within a separate 'Methods' section.
    All roxygen comment lines of a method documentation must appear after a tag.
    By default, roxygen2 will warn about any undocumented methods; if you want to have a method without documentation, use `@noRd`.

-   `@param` tags that appear before the class definition are automatically inherited by all methods, if needed.

-   R6 fields and active bindings can make use of the `@field` tag.
    Their documentation should also be in-line. To suppress documentation of a field, use `@field name NULL`.

-   roxygen2 checks that all public methods, public fields, active bindings and all method arguments are documented, and issues warnings otherwise.

-   If a method is added dynamically with `$set()`, you can document it by placing a roxygen block directly above the `$set()` call. roxygen2 will automatically associate the block with the class.

-   If roxygen2 can't automatically discover a method, you can use `@R6method Class$method` to explicitly associate a documentation block with a method. Place it in a standalone roxygen block above `NULL`:

    ```r
    #' @R6method Person$set_hair
    #' @description Change hair color.
    #' @param val New hair color.
    NULL
    ```

-   To turn off the special handling of R6 classes and go back to the roxygen2 6.x.x behavior, add `Config/roxygen2/r6: false` to your `DESCRIPTION` file.

roxygen2 automatically generates additional sections for an R6 class:

-   A section with information about the superclass(es) of the class, with links.
    In HTML this includes a list of all inherited methods, with links.

-   An 'Examples' section that contains all class and method examples.
    This section is run by `R CMD check`, so method examples must work without errors.

An example from the R6 tutorial:

```{r}
#| eval: false

#' R6 Class Representing a Person
#'
#' @description
#' A person has a name and a hair color.
#'
#' @details
#' A person can also greet you.
Person <- R6::R6Class("Person",
public = list(

    #' @field name First or full name of the person.
    name = NULL,

    #' @field hair Hair color of the person.
    hair = NULL,

    #' @description
    #' Create a new person object.
    #' @param name Name.
    #' @param hair Hair color.
    #' @return A new `Person` object.
    initialize = function(name = NA, hair = NA) {
      self$name <- name
      self$hair <- hair
      self$greet()
    },

    #' @description
    #' Change hair color.
    #' @param val New hair color.
    #' @examples
    #' P <- Person("Ann", "black")
    #' P$hair
    #' P$set_hair("red")
    #' P$hair
    set_hair = function(val) {
      self$hair <- val
    },

    #' @description
    #' Say hi.
    greet = function() {
      cat(paste0("Hello, my name is ", self$name, ".\n"))
    }
  )
)
```
