---
title: "Documenting packages"
description: >
  How to document a package as a whole using the
  `"_PACKAGE"` sentinel.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting packages}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

As well as documenting every object inside the package, you can also document the package itself by documenting the special sentinel `"_PACKAGE"`.
This automatically includes information parsed from the `DESCRIPTION`, including title, description, list of authors, and useful URLs.

We recommend placing package documentation in `{pkgname}-package.R`, and have `@keywords internal`.
Use `usethis::use_package_doc()` to set up automatically.
Here's an example:

```{r}
#| eval: false
#' @keywords internal
"_PACKAGE"
```

Package documentation is a good place to put `# Package options` that documents options used by the package.

Some notes:

- By default, aliases will be added so that both `?pkgname` and `package?pkgname` will find the package help.
  If there's an existing function called `pkgname`, use `@aliases {pkgname}-package NULL` to override the default.

- Use `@references` to point to published material about the package that users might find helpful.
