---
title: "Documenting R6"
description: >
  How to document R6 classes, methods, and fields.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting R6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Classes

Document an R6 class by placing a roxygen block before the `R6Class()` call.
Use `@description` and `@details` as usual for the class-level documentation.

roxygen2 automatically generates additional sections for an R6 class:

- A section with information about the superclass(es) of the class, with links.
  In HTML this includes a list of all inherited methods, with links.

- An 'Examples' section that contains all class and method examples.
  This section is run by `R CMD check`, so method examples must work without errors.

## Methods

R6 methods are documented **in-line**: the documentation comments go right before the definition of the method.
Method documentation can use the `@description`, `@details`, `@param`, `@return` and `@examples` tags.
These are used to create a subsection for the method, within a separate 'Methods' section.
All roxygen comment lines of a method documentation must appear after a tag.

roxygen2 checks that all public methods and their arguments are documented, and issues warnings otherwise.
If you want to have a method without documentation, use `@noRd` to suppress the warning.

`@param` tags that appear before the class definition (i.e. at the class level) are automatically inherited by all methods, if needed.

### Dynamic methods

If a method is added dynamically with `$set()`, you can document it by placing a roxygen block directly above the `$set()` call.
roxygen2 will automatically associate the block with the class.

If roxygen2 can't automatically discover a method, you can use `@R6method Class$method` to explicitly associate a documentation block with a method.
Place it in a standalone roxygen block above `NULL`:

```r
#' @R6method Person$set_hair
#' @description Change hair color.
#' @param val New hair color.
NULL
```

## Fields

R6 fields and active bindings are documented in-line using the `@field` tag.
To suppress documentation of a field, use `@field name NULL`.

roxygen2 checks that all public fields and active bindings are documented, and issues warnings otherwise.

## Example

An example from the R6 tutorial:

```{r}
#| eval: false
#' R6 Class Representing a Person
#'
#' @description
#' A person has a name and a hair color.
#'
#' @details
#' A person can also greet you.
Person <- R6::R6Class(
  "Person",
  public = list(
    #' @field name First or full name of the person.
    name = NULL,

    #' @field hair Hair color of the person.
    hair = NULL,

    #' @description
    #' Create a new person object.
    #' @param name Name.
    #' @param hair Hair color.
    #' @return A new `Person` object.
    initialize = function(name = NA, hair = NA) {
      self$name <- name
      self$hair <- hair
      self$greet()
    },

    #' @description
    #' Change hair color.
    #' @param val New hair color.
    #' @examples
    #' P <- Person("Ann", "black")
    #' P$hair
    #' P$set_hair("red")
    #' P$hair
    set_hair = function(val) {
      self$hair <- val
    },

    #' @description
    #' Say hi.
    greet = function() {
      cat(paste0("Hello, my name is ", self$name, ".\n"))
    }
  )
)
```

## Opting out

To turn off the special handling of R6 classes and go back to the roxygen2 6.x.x behavior, add `Config/roxygen2/r6: false` to your `DESCRIPTION` file.
