---
title: "Documenting R6"
description: >
  How to document R6 classes, methods, and fields.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting R6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are three things that you might document for [R6](https://adv-r.hadley.nz/r6.html):

- **Classes**: document the class as a whole with a roxygen block before `R6Class()`.
- **Fields**: document fields and active bindings in-line with `@field`.
- **Methods**: document methods in-line; every method doc must start with a tag.

R6 documentation works differently from regular function documentation because much of it is written **in-line**, inside the class definition.
Additionally, because `R CMD check` doesn't know about R6, roxygen2 is in charge of checking that all public methods, fields, active bindings, and method arguments are documented.
It'll issue warnings for anything that's missing, which you can suppress using the techniques described below.

## Classes

Document an R6 class by placing a roxygen block before the `R6Class()` call.
Use `@export` to make the class available to users.

```{r}
#| eval: false
#' R6 Class Representing a Person
#'
#' A person has a name and a hair color.
#' @export
Person <- R6::R6Class(
  "Person",
  public = list(
    # ... fields and methods ...
  )
)
```

roxygen2 automatically generates additional sections:

- A section with information about the superclass(es) of the class, with links.
  In HTML this includes a list of all inherited methods, with links.

- An 'Examples' section that contains all class and method examples.
  This section is run by `R CMD check`, so method examples must work without errors.

`@param` tags that appear at the class level (i.e. before the `R6Class()` call) are automatically inherited by all methods, if needed.

## Fields and active bindings

Fields and active bindings are documented **in-line** using the `@field` tag, right before the field definition:

```{r}
#| eval: false
Person <- R6::R6Class(
  "Person",
  public = list(
    #' @field name First or full name of the person.
    name = NULL,

    #' @field birthdate Date of birth, as a [Date].
    birthdate = NULL
  ),
  active = list(
    #' @field age Age in years, computed from `birthdate` (read-only).
    age = function() {
      as.numeric(difftime(Sys.Date(), self$birthdate, units = "days")) / 365.25
    }
  )
)
```

To suppress documentation of a field or active binding, use `@field name NULL`.

## Methods

Methods are also documented **in-line**, right before the method definition.
Unlike regular function documentation, all roxygen comment lines for a method must appear **after a tag** --- you can't start with a plain text introduction.
Use `@description` to provide the method's description:

```{r}
#| eval: false
Person <- R6::R6Class(
  "Person",
  public = list(
    #' @description
    #' Create a new person object.
    #' @param name Name.
    #' @param hair Hair color.
    #' @returns A new `Person` object.
    initialize = function(name = NA, hair = NA) {
      self$name <- name
      self$hair <- hair
      self$greet()
    },

    #' @description
    #' Change hair color.
    #' @param val New hair color.
    set_hair = function(val) {
      self$hair <- val
    },

    #' @description
    #' Say hi.
    greet = function() {
      cat(paste0("Hello, my name is ", self$name, ".\n"))
    }
  )
)
```

Like functions, methods can use the `@description`, `@details`, `@param`, `@returns`, and `@examples` tags.
These are used to create a subsection for the method, within a separate 'Methods' section in the rendered help.

If you want to leave a method without documentation, use `@noRd` to suppress the warning.

### Dynamic methods

If a method is added dynamically with `$set()`, you can document it by placing a roxygen block directly above the `$set()` call:

```{r}
#| eval: false
#' @description
#' Say goodbye.
Person$set("public", "goodbye", function() {
  cat(paste0("Goodbye from ", self$name, ".\n"))
})
```

roxygen2 will automatically associate the block with the class.

If roxygen2 can't automatically discover a method, you can use `@R6method Class$method` to explicitly associate a documentation block with a method.
Place it in a standalone roxygen block above `NULL`:

```r
#' @R6method Person$set_hair
#' @description Change hair color.
#' @param val New hair color.
NULL
```

## Opting out

To turn off the special handling of R6 classes and go back to the roxygen2 6.x.x behavior, add `Config/roxygen2/r6: false` to your `DESCRIPTION` file.
