---
title: "Documenting S4"
description: >
  How to document S4 generics, methods, and classes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting S4}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

There are three things that you might document for [S4](https://adv-r.hadley.nz/s4.html):

- **Generics**: document as functions; mention that they are generics.
- **Methods**: unlike S3/S7, all S4 methods must be documented.
- **Classes**: document the class definition and use `@slot` for slots.

## Generics

S4 **generics** are functions, so document them as such.
`@export` a generic if you want users to call it or other developers to write methods for it.
If the generic is internal, you don't need to export or document it.

## Classes

Document **S4 classes** by adding a roxygen block before `setClass()`.
`@export` a class if you want users to create instances or other developers to extend it (e.g. by creating subclasses).
Internal classes don't need documentation.
Use `@slot` to document the slots of the class.
Here's a simple example:

```{r}
#| eval: false
#' An S4 class to represent a bank account
#'
#' @slot balance A length-one numeric vector
#' @export
Account <- setClass("Account", slots = list(balance = "numeric"))
```

## Methods

S4 **methods** are a little more complicated.
Unlike S3 and S7 methods, all S4 methods must be documented.
You only need to `@export` a method if the generic lives in another package.

You can document methods in three places:

- In the class.
  Most appropriate if the corresponding generic uses single dispatch and you created the class.

- In the generic.
  Most appropriate if the generic uses multiple dispatches and you control it.

- In its own file.
  Most appropriate if the method is complex or the either two options don't apply.

Use either `@rdname` or `@describeIn` to control where method documentation goes.
See `vignette("reuse")` for more details.
