---
title: "Documenting S3"
description: >
  How to document S3 generics, methods, and classes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting S3}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Documenting [S3 code](https://adv-r.hadley.nz/s3.html) well requires thinking about three things:

- **Generics**: mention that the function is a generic and list the available methods.
- **Methods**: link back to the generic; only document individually when the method has unique behavior or arguments.
- **Classes**: document the constructor.

## Generics

S3 **generics** are regular functions, so document them as such.
Export a generic if you want users to call it or other developers to write methods for it.
If the generic is internal, you don't need to document it.

The documentation should mention that the function is a generic, because this tells the reader that the behavior may vary depending on the input and that they can write their own methods.
For simple generics, you can do this in the description:

```{r}
#| eval: false
#' Frobnpolicate an object
#'
#' @description
#' `frobnpolicate()` is an S3 generic that ..., with methods available for
#' the following classes:
#'
#' `r doclisting::methods_inline("frobnpolicate")`
```

For more complicated generics, you can use a `# Methods` section to provide more detail:

```{r}
#| eval: false
#' Frobnpolicate an object
#'
#' @description
#' `frobnpolicate()` does ...
#'
#' # Methods
#' `frobnpolicate()` is an S3 generic with methods available for the following
#' classes:
#'
#' `r doclisting::methods_list("frobnpolicate")`
```

You might also want to include a section that provides additional details for developers implementing their own methods.

Both examples above use the [doclisting](https://doclisting.r-lib.org/) package to automatically generate a list of methods, with links to their help topics.
These examples use [inline R code](reuse.html#inline-code) (`` `r ` ``), which generates the list at documentation time (i.e.
when you run `devtools::document()`).
This only requires including doclisting in `Suggests`.

If you want the list to dynamically reflect all methods that are currently registered (including methods registered by other packages), use an [inline Rd code](reuse.html#inline-rd-code) block (`` `Rd ` ``) instead:

```{r}
#| eval: false
#' `Rd doclisting::methods_list("frobnpolicate")`
```

Using `` `Rd ` `` requires doclisting in `Imports`, and you'll need a dummy call to eliminate the `R CMD check` NOTE about unused imports:

```{r}
#| eval: false
ignore_unused_imports <- function() {
  doclisting::methods_list
}
```

## Classes

S3 **classes** have no formal definition, so document the [constructor](https://adv-r.hadley.nz/s3.html#s3-constructor).
Export the constructor if you want users to create instances of your class or other developers to extend it (e.g.
by creating subclasses).
Internal constructors don't need documentation.

Note that you don't need to list methods for a class: in S3, methods belong to generics, not to classes.
Users should look at the generic's help page to learn about available methods.

## Methods

It is your choice whether or not to document S3 **methods**.
Generally, it's not necessary to document straightforward methods for common generics like `print()`.
(You should, however, always `@export` S3 methods, even internal ones.)

It's good practice, however, to document methods that have unique behavior or arguments.
For example, the clock package documents the methods for [`date_group()`](https://clock.r-lib.org/reference/date_group.html) on their own pages ([Date](https://clock.r-lib.org/reference/date-group.html), [POSIXt](https://clock.r-lib.org/reference/posixt-group.html)) because the methods accept different `precision` values and have type-specific return value semantics and examples.
The generic page serves only as a signpost linking to the individual method pages.

When documenting a method, always include a link back to the generic using `[generic_name()]` so the reader can easily find the full documentation and other methods.

## Method disambiguation

Generally, roxygen2 will automatically figure out the generic that the method belongs to, and you should only need to use `@method` if there is ambiguity.
For example, is `all.equal.data.frame()` the `equal.data.frame` method for `all()`, or the `data.frame` method for `all.equal()`?
If this happens to you, disambiguate with (e.g.)
`@method all.equal data.frame`.
