<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Generating Rd files}
-->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE
)
```

## Basic documentation

Roxygen comments have a special format. They start with `#'` and include tags that like `@tag details`. Tags break the documentation up into pieces, and the content of a tag extends from the end of tag name to the start of the next tag (or the end of the block). Because `@` has a special meaning in roxygen, you need to write `@@` to add a literal `@` to the documentation.

Each documentation block starts with some text. The first sentence becomes the title of the documentation. That's what you see when you look at `help(package = mypackage)` and is shown at the top of each help file. It should fit on one line, be written in sentence case, and end in a full stop. The second paragraph is the description: this comes first in the documentation and should briefly describe what the function does. The third and subsequent paragraphs go into the details: this is a (often long) section that comes after the argument description and should provide any other important details of how the function operates.

Here's an example showing what the documentation for `sum()` might look like if it had been written with roxygen:

```{r}
#' Sum of vector elements.
#'
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

`\code{}` and `\link{}` are `.Rd` formatting commands which you'll learn more about in [formatting](formatting.html). Also notice the wrapping of the roxygen block. You should make sure that your comments are less than ~80 columns wide.

The following documentation produces the same help file as above, but uses explicit tags. You only need explicit tags if you want to the title or description to span multiple paragraphs (a bad idea), or want to omit the description (in which case roxygen will use the title for the description, since it's a required documentation component).

```{r}
#' @title Sum of vector elements.
#'
#' @description
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' @details
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

All objects must have a title and description. Details are optional.

There are two tags that make it easier for the user to navigate around your document. `@seealso` allows you to point to other useful resources, either on the web `\url{http://www.r-project.org}`, or to other documentation with `\code{\link{functioname}}`. If you have a family of related functions that all needed to be interlinked, you can use the `@family` tag to automatically add the appropriate links to seealso. The `@family` name should be plural.

For sum, these components might look like:

```{r}
#' @family aggregate functions
#' @seealso \code{\link{prod}} for products, \code{\link{cumsum}} for
#'  cumulative sums, and \code{\link{colSums}}/\code{\link{rowSums}}
#'  marginal sums over high-dimensional arrays.
```

You use other tags based on the type of object that you're documenting. The following sections describe the most commonly used tags for functions, S3, S4 and RC objects and data.

## Documenting functions

Functions are the mostly commonly documented objects. Most functions use three tags:

* `@param name description` describes the inputs to the function.
  The description should provide a succinct summary of the type of the
  parameter (e.g. a string, a numeric vector), and if not obvious from
  the name, what the parameter does. The description can span multiple
  lines (or even paragraphs) if necessary. All parameters must be
  documented.

  It's possible to document multiple arguments in one place by separating
  the names with commas (no spaces). For example, to document both
  `x` and `y`, you can say `@param x,y Numeric vectors.`.

* `@examples` provides executable R code showing how to use the function in
  practice. This is a very important part of the documentation because
  many people look at the examples before reading anything. Example code
  must work without errors as it is run automatically as part of `R CMD
  check`.

    For the purpose of illustration, it's often useful to include code
    that causes an error. `\dontrun{}` allows you to include code in the
    example that is never use. There are two other special commands
    Code `\dontshow{}` is run, but not shown in the help page: this can
    be useful for informal tests. `\donttest{}` is run in examples,
    but not run automatically but `R CMD check`. This is useful if you
    have examples that take a long time to run. The options are summarised
    below.

    Command      | example   | help   | R CMD check
    ------------ | --------- | ------ | -----------
    `\dontrun{}` |           | x      |
    `\dontshow{}`| x         |        | x
    `\donttest{}`| x         | x      | x

* `@return description` describes the output from the function. This is
  not always necessary, but is a good idea if you return different types
  of outputs depending on the input, or you're returning an S3, S4 or RC
  object.

We could use these new tags to improve our documentation of `sum()` as follows:

```{r}
#' Sum of vector elements.
#'
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
#'
#' @param ... Numeric, complex, or logical vectors.
#' @param na.rm A logical scalar. Should missing values (including NaN)
#'   be removed?
#' @value If all inputs are integer and logical, then the output
#'   will be an integer. If integer overflow
#'   \url{http://en.wikipedia.org/wiki/Integer_overflow} occurs, the output
#'   will be NA with a warning. Otherwise it will be a length-one numeric or
#'   complex vector.
#'
#'   Zero-length vectors have sum 0 by definition. See
#'   \url{http://en.wikipedia.org/wiki/Empty_sum} for more details.
#' @examples
#' sum(1:10)
#' sum(1:5, 6:10)
#' sum(F, F, F, T, T)
#'
#' sum(.Machine$integer.max, 1L)
#' sum(.Machine$integer.max, 1)
#'
#' \dontrun{
#' sum("a")
#' }
sum <- function(..., na.rm = TRUE) {}
```

Indent the second and subsequent lines of a tag so that when scanning the documentation it's easy to see where one tag ends and the next begins. Tags that always span multiple lines (like `@example`) should start on a new line and don't need to be indented.

## Documenting multiple functions in the same file

You can document multiple functions in the same file by using the `@rdname` tag. This overrides the default file name generated by roxygen and merges documentation for mulitple objects into one file. It's a technique best used with caution: documenting too many functions into one place leads to confusing documentation. It's best used when all functions have the same (or very similar) arguments.

There are two ways to use `@rdname`. You can add documentation to an existing function:

```{r}
#' Basic arithmetic
#'
#' @param x,y numeric vectors.
add <- function(x, y) x + y

#' @rdname add
times <- function(x, y) x * y
```

Or, you can create a dummy documentation file by documenting `NULL` and setting an informative `@name`.

```{r}
#' Basic arithmetic
#'
#' @param x,y numeric vectors.
#' @name arith
NULL

#' @rdname arith
add <- function(x, y) x + y

#' @rdname arith
times <- function(x, y) x * y
```

## Documenting classes, generics and methods

Documenting classes, generics and methods are relatively straightforward, but there are some variations depending on which object system you're using. The following sections give the details for the S3, S4 and RC object systems.

### S3

S3 __generics__ are regular functions, so document them as such. S3 __classes__ have no formal definition, so document the constructor function. It is your choice whether or not to document S3 __methods__. You don't need to document methods for simple generics like `print()`. If your method is more complicated, you should document it so people know what the parameters do. In base R, you can find documentation for more complex methods like `predict.lm()`, `predict.glm()`, and `anova.glm()`.

Older versions of roxygen required explicit `@method generic class` tags for all S3 methods. From 3.0.0 this is no longer needed as and roxygen2 will figure it out automatically. If you are upgrading, make sure to remove these old tags. Automatic method detection will only fail if the generic and class are ambiguous. For example is `all.equal.data.frame()` the `equal.data.frame` method for `all`, or the `data.frame` method for `all.equal`?. If this happens, you can disambiguate with (e.g.) `@method all.equal data.frame`.

### S4

Older versions of roxyen2 required explicit `@usage`, `@alias` and `@docType` to correctly document S4 objects, but from version 3.0.0 on roxygen2 generates correct metadata automatically. If you're upgrading from a previous version, make sure to remove these old tags.

S4 __generics__ are also functions, so document them as such. Document __S4 classess__ by adding a roxygen block before `setClass()`. Use `@slot` to document the slots of the class. Here's a simple example:

```{r}
#' An S4 class to represent a bank account.
#'
#' @slot balance A length-one numeric vector
Account <- setClass("Account",
  slots = list(balance = "numeric")
)
```

S4 __methods__ are a little more complicated. Unlike S3, all S4 methods must be documented. You can document them in three places:

* In the class. Most appropriate if the corresponding generic uses single
  dispatch and you created the class.

* In the generic. Most appropriate if the generic uses multiple dispatch
  and you control it.

* In it's own file. Most appropriate if the method is complex. or the
  either two options don't apply.

Use `@rdname` to put the method documentation where you want it.

### RC

RC is different to S3 and S4 because methods are associated with classes, not generics. RC also has a special convention for documenting methods: the doc string. This makes documenting RC simpler than S4 because you only need one  roxygen block per class.

```{r}
#' An S4 class to represent a bank account.
#'
#' @field balance A length-one numeric vector
Account <- setRefClass("Account",
  fields = list(balance = "numeric"),
  methods = list(
    withdraw <- function(x) {
      "Withdraw money from account. Allows overdrafts"
      balance <<- balance - x
    }
  )
)
```

Methods with doc strings will be included in the "Methods" section of the class documentation. Each documented method will be listed with an automatically generated usage statement and its doc string.

## Documenting datasets

Datasets are usually stored as `.rdata` files in `data/` and not as regular R objects in the package. This means you need document them slightly differently: instead of documenting the data directly, you document `NULL`, and use `@name` to tell roxygen2 what dataset you're really documenting.

There are two new tags that are useful for documenting datasets:

* `@format`, which gives an overview of the structure of the dataset.
  If you omit this, roxygen will automatically add something based on the
  first line of `str()` output

* `@source` where you got the data form, often a `\url{}`.

To show how everything fits together, the example below is an exercept from the roxygen block used to document the diamonds dataset in ggplot2.

```{r}
#' Prices of 50,000 round cut diamonds.
#'
#' A dataset containing the prices and other attributes of almost 54,000
#' diamonds. The variables are as follows:
#'
#' \itemize{
#'   \item price. price in US dollars (\$326--\$18,823)
#'   \item carat. weight of the diamond (0.2--5.01)
#'   ...
#' }
#'
#' @format A data frame with 53940 rows and 10 variables
#' @source \url{http://www.diamondse.info/}
#' @name diamonds
NULL
```

## Documenting packages

As well as documenting every exported object in the package, you should also document the package itself. Relatively few packages provide package documentation, but it's an extremely useful tool for users, because instead of just listing functions like `help(package = pkgname)` it organises them and shows the user where to get started.

Package documentation should describe the overall purpose of the package and point to the most important functions. It should not contain a verbatim list of functions or copy of `DESCRIPTION`. This file is for human reading, so pick the most important elements of your package.

Package documentation should be placed in `pkgname.R`. Here's an example:

```{r}
#' Generate R documentation from inline comments
#'
#' Roxygen2 allows you to write documentation in comment blocks co-located
#' with code.
#'
#' The only function you're likely to need from \pkg{roxygen2} is
#' \code{\link{roxygenize}}. Otherwise refer to the vignettes to see
#' how to format the documentation.
#'
#' @docType package
#' @name roxygen2
NULL
```

Some notes:

* Like for datasets, there isn't a object that we can document directly so
  document `NULL` and use `@name` to say what we're actually documenting

* `@docType package` indicates that the documentation is for the package.
  This will automatically add the corect aliases so that both `?pkgname`
  and `package?pkgname` will find the package help. If there's already
  a function called `pkgname()`, use `@name roxygen2-package`.

* Use `@references` point to published material about the package that
  users might find helpful.

## Sections

For any object, you can add arbitrary sections to the documentation with the `@section` tag. This is a useful way of breaking a long details section into multiple chunks with useful headings. Section titles should be in sentence case and must be followed a colon. Titles may only take one line.

```{r}
#' @section Warning:
#' Do not operate heavy machinery within 8 hours of using this function.
```

To add a subsection, you must use the `Rd` `\subsection{}` command, as follows:

```{r}
#' @section Warning:
#' You must not call this function unless ...
#'
#' \subsection{Exceptions}{
#'    Apart from the following special cases...
#' }
```
