<!--
%\VignetteEngine{knitr}
%\VignetteIndexEntry{Generating Rd files}
-->

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(
  comment = "#>",
  error = FALSE,
  tidy = FALSE
)
```

## Basic documentation

As you've seen, roxygen comments have a special format. They start with `#'` and include tags that like `@tag details`. Tags break the documentation up into pieces, and the content of a tag extends from the end of tag name to the start of the next tag (or the end of the block). Because `@` has a special meaning in roxygen, you need to write `@@` to add a literal `@` to the documentation.

Each documentation block starts with some text. The first sentence becomes the title of the documentation. That's what you see when you look at `help(package = mypackage)` and is shown at the top of each help file. It should fit on one line, be written in sentence case, and end in a full stop. The second paragraph is the description: this comes first in the documentation and should briefly describe what the function does. The third and subsequent paragraphs go into the details: this is a (often long) section that comes after the argument description and should provide any other important details of how the function operates.

Here's an example showing what the documentation for `sum()` might look like if it had been written with roxygen:

```{r}
#' Sum of vector elements.
#'
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

`\code{}` and `\link{}` are `.Rd` formatting commands. Learn more about them and other tools in [formatting](formatting.html). Also notice the wrapping of the roxygen block. You should make sure that your comments are less than ~80 columns wide.

The following documentation is equivalent, but uses explicit tags. This is not necessary unless you want to have a multiple paragraph description or title, or want to omit the description (in which case roxygen will use the title for the description, since it's a required component of the `.Rd` file). None of these actions are recommended so you have to go out of your way to get them.

```{r}
#' @title Sum of vector elements.
#'
#' @description
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' @details
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

All objects must have a title, description and details. Objects can also have `@author`, in the form `@author Firstname Lastname <email@@address.com>`. You only need to use this if the function was written by someone author than an author already listed in the DESCRIPTION.

There are two tags that make it easier for the user to navigate around your document. `@seealso` allows you to point to other useful resources, either on the web `\url{http://www.r-project.org}`, or to other document with `\code{\link{functioname}}`. If you have a family of related functions that all needed to be interlinked, you can use the `@family` tag to automatically add seealso's. The `@family` name should be plural.

For sum, these components might look like:

```{r}
#' @family aggregate functions
#' @seealso \code{\link{prod}} for products, \code{\link{cumsum}} for
#'  cumulative sums, and \code{\link{colSums}}/\code{\link{rowSums}}
#'  marginal sums over high-dimensional arrays.
```

You use other tags based on the type of object that you're documenting. The following sections describe the most commonly used tags for functions, S3, S4 and RC objects and data.

## Documenting functions

Functions are the mostly commonly documented objects. Most functions require three new tags:

* `@param name description` describes the inputs to the function.
  The description should provide a succinct summary of the type of the
  parameter (e.g. a string, a numeric vector), and if not obvious from
  the name, what the parameter does. The description can span multiple
  lines (or even paragraphs) if necessary. All parameters must be
  documented.

  It's possible to document multiple arguments at once by separating
  the names with commas (and no whitespace). For example, to document both
  `x` and `y`, you can say `@param x,y Numeric vectors`.

* `@examples` provides examples of the function in use. This is a very
  important part of the documentation. Many people look at the examples
  before reading anything else in documentation.

* `@return description` describes the output from the function. This is
  not always necessary, but is a good idea if you return different types
  of outputs depending on the input, or you're returning an S3, S4 or RC
  object.

Using these elements for the documentation of `sum()` might yield something like this:

```{r}
#' Sum of vector elements.
#'
#' \code{sum} returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the \code{\link{Summary}} group generic. For this to work properly,
#' the arguments \code{...} should be unnamed, and dispatch is on the
#' first argument.
#'
#' @param ... Numeric, complex, or logical vectors.
#' @param na.rm A logical scalar. Should missing values (including NaN)
#'   be removed?
#' @value If all inputs are integer and logical, then the output
#'   will be an integer. If integer overflow
#'   \url{http://en.wikipedia.org/wiki/Integer_overflow} occurs, the output
#'   will be NA with a warning. Otherwise it will be a length-one numeric or
#'   complex vector.
#'
#'   Zero-length vectors have sum 0 by definition. See
#'   \url{http://en.wikipedia.org/wiki/Empty_sum} for more details.
#' @examples
#' sum(1:10)
#' sum(1:5, 6:10)
#' sum(F, F, F, T, T)
#'
#' sum(.Machine$integer.max, 1L)
#' sum(.Machine$integer.max, 1)
sum <- function(..., na.rm = TRUE) {}
```

Indent the second and subsequent lines of a tag so that when scanning the documentation it's easy to see where one tag ends and the next begins. Tags that always span multiple lines (like `@example`) should start on a new line and don't need to be indented.

## Documenting multiple functions in the same file

It's possible to document multiple functions in the same file by using the `@rdname` tag. This overrides the default file name generated by roxygen and merges documentation for mulitple objects into one file. It's a technique best used with caution: documenting too many functions into one place leads to confusing documentation. It's best used when all functions have the same (or very similar) arguments.

There are basically two ways to use it. First, you can combine documentation into an existing function

```{r}
#' This is the main function
#'
#' @param x,y,z numeric vectors.
f <- function(x, y, z) {}

#' @rdname f
g <- function(x, y, z) {}
```

Or create a dummy documentation file:

```{r}

#' This is the main function
#'
#' @param x,y,z numeric vectors.
#' @name my-object
NULL

#' @rdname my-object
f <- function(x, y, z) {}

#' @rdname my-object
g <- function(x, y, z) {}
```

## S3

### S3 generics

Documenting S3 generics is straightforward: they are functions, so document them as such.

### S3 classes

S3 classes have no formal definition, so document the constructor function.

### S3 methods

It is your choice whether or not to document S3 methods. You don't need to document methods for simple generics like `print()`. If your method is more complicated, you should document it so people know what the parameters do. In base R, you can find documentation for more complex methods like `predict.lm()`, `predict.glm()`, and `anova.glm()`.

Older versions of roxygen required explicit `@method generic class` tags for all S3 methods. From 3.0.0 this is no longer needed as and roxygen2 will figure it out automatically. If you are upgrading, make sure to remove these old tags. Automatic method detection will only fail if the generic and class are ambiguous. For example is `all.equal.data.frame()` the `equal.data.frame` method for `all`, or the `data.frame` method for `all.equal`?. If this happens, you can disambiguate with (e.g.) `@method all.equal data.frame`.

## S4

Older versions of roxyen2 required explicit `@usage`, `@alias` and `@docType` to correctly document S4 objects, but from version 3.0.0 on roxygen2 generates correct metadata automatically. If you're upgrading from a previous version, make sure to remove these old tags.

### S4 generics

S4 generics are also functions, so document them as such.

### S4 classes

To document an S4 class, add a roxygen block before `setClass()`. Use `@slot` to document the slots of the class. Here's a simple example:

```{r}
#' An S4 class to represent a bank account.
#'
#' @slot balance A length-one numeric vector
Account <- setClass("Account",
  slots = list(balance = "numeric")
)
```

### S4 methods

Unlike S3, all S4 methods must be documented. You can document them in three places:

* In the class. Most appropriate if the corresponding generic uses single
  dispatch and you created the class.

* In the generic. Most appropriate if the generic uses multiple dispatch
  and you control it.

* In it's own file. Most appropriate if the method is complex. or the
  either two options don't apply.

Use `@rdname` to put the method documentation where you want it.

## RC

RC is different to S3 and S4 because it does not have generics and methods are associated with classes, not generics. RC also has a different convention for documenting methods: the doc string. This makes documenting RC classes a little simpler than S4 classes, generics and methods because you only need a single roxygen block per class.

```{r}
#' An S4 class to represent a bank account.
#'
#' @field balance A length-one numeric vector
Account <- setRefClass("Account",
  fields = list(balance = "numeric"),
  methods = list(
    withdraw <- function(x) {
      "Withdraw money from account. Allows overdrafts"
      balance <<- balance - x
    }
  )
)
```

Documented methods will be included in the "Methods" section of the class documentation, along with an automatically generated usage statement.

## Documenting datasets

Datasets are usually stored in `data/` and not as regular R objects in the package. This means you need document them slightly differently: instead of documenting the data directly, you document `NULL`, and use `@name` to tell roxygen2 what you're really documenting.

There are two new tags that are useful for documenting datasets:

* `@format`, which gives an overview of the structure of the dataset. If you omit this, roxygen will automatically add something based on the first line of `str` output

* `@source` where you got the data form, often a `\url{}`.

To show how everything fits together, the example below is an exercept from the roxygen block using to document the diamonds dataset in ggplot2.

```{r}
#' Prices of 50,000 round cut diamonds.
#'
#' A dataset containing the prices and other attributes of almost 54,000
#' diamonds. The variables are as follows:
#'
#' \itemize{
#'   \item price. price in US dollars (\$326--\$18,823)
#'   \item carat. weight of the diamond (0.2--5.01)
#'   ...
#' }
#'
#' @format A data frame with 53940 rows and 10 variables
#' @source \url{http://www.diamondse.info/}
#' @name diamonds
NULL
```

## Documenting packages

As well as documenting every exported object in the package, you should also document the package itself. Package documentation should describe the overall purpose of the package and point to the most important functions. Relatively few packages provide package documentation, but it's an extremely useful tool for users, because instead of just listing functions like `help(package = pkgname)` it organises them and shows the user where to get started.

I usually put this documentation in a `.R` file with the same name as the package.

* Like for data sets, there isn't a object that we can document directly so
  document `NULL` and use `@name` to say what we're actually documenting

* `@docType package` to indicate that it's documenting a package. This will
  automatically add the corect aliases so that `package?yourpackage` works.

* `@references` point to any published material about the package that users
  might find help.

* The topic will be automatically aliases to `package-pkgname` and `pkgname` (unless there is already a function of that name) so that you can find the package help with both `?pkgname` and `package?pkgname`.

The package documentation should not contain a verbatim list of functions or copy of `DESCRIPTION`. This file is for human reading, so pick the most important elements of your package.


## Sections

You can add arbitrary sections to the documentation with the `@section` tag. The section title should be in sentence case and must be followed a colon. Titles may only take one line.

```{r}
#' @section Warning:
#' Do not operate heavy machinery within 8 hours of using this function.
```

To add a subsection, you must use the `Rd` `\subsection{}` command, as follows:

```{r}
#' @section Warning:
#' You must not call this function unless ...
#'
#' \subsection{Exceptions}{
#'    Apart from the following special cases...
#' }
```
