---
title: "Documenting functions"
description: >
  The basics of roxygen2 tags and how to use them for 
  documenting functions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documenting functions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(comment = "#>", collapse = TRUE)
```

This vignette introduces the basics of roxygen2 for documenting functions.
See `vignette("rd-other")` for documenting other types of objects and `vignette("reuse")` for reusing documentation across topics.

## Basics

A roxygen **block** is a sequence of lines starting with `#'` (optionally preceded by white space).

The first lines of the block is called the **introduction** and forms the title, description, and details, as described below.
The introduction continues until the first **tag**.

Tags start with `@`, like `@details` or `@param`.
Tags must appear at the beginning of a line and their content extends to the start of the next tag or the end of the block.
Text within the description or tags can be formatted using Markdown or `Rd` commands; see `vignette("rd-formatting")` for details.

A block ends when it hits R code, usually a function or object assignment.
Blocks ignore empty lines, including lines made up of non-roxygen comments.
If you need to separate two blocks, use `NULL`.

## The introduction

Each documentation block starts with some text which defines the title, the description, and the details.
Here's an example showing what the documentation for `sum()` might look like if it had been written with roxygen:

```{r}
#' Sum of vector elements
#'
#' `sum` returns the sum of all the values present in its arguments.
#'
#' This is a generic function: methods can be defined for it directly
#' or via the [Summary()] group generic. For this to work properly,
#' the arguments `...` should be unnamed, and dispatch is on the
#' first argument.
sum <- function(..., na.rm = TRUE) {}
```

This introductory block is broken up as follows:

-   The first sentence is the **title**: that's what you see when you look at `help(package = mypackage)` and is shown at the top of each help file.
    It should generally fit on one line, be written in sentence case, and not end in a full stop.

-   The second paragraph is the **description**: this comes first in the documentation and should briefly describe what the function does.

-   The third and subsequent paragraphs go into the **details**: this is a (often long) section that comes after the argument description and should provide any other important details of how the function operates.
    The details are optional.

You can also use explicit `@title`, `@description`, and `@details` tags.
This is unnecessary unless you want to have a multi-paragraph description, bulleted list, or other more exotic structure.

```{r}
#' Sum of vector elements
#' 
#' @description
#' `sum` returns the sum of all the values present in its arguments.
#'
#' @details
#' This is a generic function: methods can be defined for it directly
#' or via the [Summary()] group generic. For this to work properly,
#' the arguments `...` should be unnamed, and dispatch is on the
#' first argument.
```

## Functions

Functions are the mostly commonly documented objects.
Functions require three tags: `@param`, `@returns`, and `@examples`.

### Inputs

Use `@param name description` to describe each input to the function.
The description should provide a succinct summary of parameter type (e.g. a string, a numeric vector), and if not obvious from the name, what the parameter does.
The description is a sentence so should start with a capital letter and end with a full stop.
It can span multiple lines (or even paragraphs) if necessary.
All parameters must be documented.

If two or more arguments are tightly coupled, you can document them in one place by separating the names with commas (no spaces).
For example, to document both `x` and `y`, you can say `@param x,y Numeric vectors`.

### Outputs

`@returns description` describes the output from the function.
Briefly describe the type/shape of the output, not the details.

All functions must have a documented return value for initial CRAN submission.

### Examples

`@examples` provides executable R code showing how to use the function in practice.
This is a very important part of the documentation because many people look at the examples before reading anything.
Example code must work without errors as it is run automatically as part of `R CMD check`.

For the purpose of illustration, it's often useful to include code that causes an error.
You can do this by wrapping the code in `try()` or using `\dontrun{}` to exclude from the executed example code.

For finer control, you can use `@examplesIf`:

``` r
#' @examplesIf interactive()
#' browseURL("https://roxygen2.r-lib.org")
```

This generates

    \examples{
    \dontshow{if (interactive() (if (getRversion() >= "3.4") withAutoprint else force)(\{ # examplesIf}
    gh_organizations(since = 42)
    \dontshow{\}) # examplesIf}
    }

This way, the code evaluating whether the example can be run is not shown to users reading the help, but it still prevents R CMD check failures.

Instead of including examples directly in the documentation, you can put them in separate files and use `@example path/relative/to/package/root` to insert them into the documentation.

All functions must have a documented return value for initial CRAN submission.

### Usage

In most case, the function usage (which appears beneath the description in the generates docs) will be automatically derived from the function specification.
For the cases where it is not, please [file an issue](https://github.com/r-lib/roxygen2/issues) and use `@usage` to override the default with you want.
If you want to suppress the usage altogether (which is sometimes useful for internal or deprecated functions), you can use `@usage NULL`.

## Other tags

### Cross-references

There are two tags that make it easier for people to navigate your documentation: `@seealso` and `@family`.
`@seealso` allows you to point to other useful resources, either on the web `<http://www.r-project.org>` or to other documentation with `[function_name()]`.
If you have a family of related functions, you can use `@family {family}` to cross-reference each function to every other function within the family.
A function can be a member of multiple families.

For `sum()`, this might look like:

```{r}
#' @family aggregations
#' @seealso [prod()] for products, [cumsum()] for cumulative sums, and
#'   [colSums()]/[rowSums()] marginal sums over high-dimensional arrays.
```

By default `@family {family}`, will generate the see also text "Other {family}:", so the `@family` name should be plural (i.e., "model building helpers" not "model building helper").
You can override the default title by providing a `rd_family_title` list in `man/roxygen/meta.R`:

```{r, eval = FALSE}
list(
  rd_family_title = list(aggregations = "Aggregation functions")
)
```

### Suppress documentation

It's sometimes useful to include roxygen2 documentation tags without generating an `.Rd` file.
You can use `@noRd` to suppress file generation for a given topic.

### Indexing

Three other tags make it easier for the user to find documentation within R's help system:

-   Aliases form the index that `?` searches.
    Use `@aliases space separated aliases` to add additional aliases.

-   `@concept` add extra keywords that will be found with `help.search()`

-   Use `@keywords keyword1 keyword2 ...` to add standardised keywords.
    Keywords are optional, but if present, must be taken from the predefined list found `file.path(R.home("doc"), "KEYWORDS")`.

Apart from `@keywords internal`, these tags are not very useful because most people find documentation using Google.
`@keywords internal` is useful because it removes the function from the documentation index; it's useful for functions aimed primarily at other developers, not typical users of the package.

### Back references

The original source location is added as a comment to the second line of each generated `.Rd` file in the following form:

    % Please edit documentation in ...

`roxygen2` tries to capture all locations from which the documentation is assembled.
For code that *generates* R code with Roxygen comments (e.g., the Rcpp package), the `@backref` tag is provided.
This allows specifying the "true" source of the documentation, and will substitute the default list of source files.
Use one tag per source file:

```{r}
#' @backref src/file.cpp
#' @backref src/file.h
```

