---
title: "Reusing documentation"
description: >
  Tools for reusing documentation across topics, and between
  documentation and vignettes.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reusing documentation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

roxygen2 provides several ways to avoid repeating yourself in code documentation, while assembling information from multiple places in one documentation file:

-   Combine documentation for closely related functions into a single file with `@describe` in or `@rdname`.

-   Inherit components from another topic with `@inheritParams`, `@inheritSection`, or `@inherit`.

-   Use functions to generate repeated text with inline R code.

-   Share text between documentation and vignettes with child documents.

The chapter concludes by showing you how to update superseded reuse mechanisms that we no longer recommend: `@includeRmd`, `@eval`/`@evalRd`, and `@template`.

## Multiple functions in the same topic

You can document multiple functions in the same file by using either `@rdname` or `@describeIn` tag.
It's a technique best used with care: documenting too many functions in one place leads to confusion.
Use it when all functions have the same (or very similar) arguments.

### `@describeIn`

`@describeIn` is designed for the most common cases:

-   documenting methods in a generic
-   documenting methods in a class
-   documenting functions with the same (or similar arguments)

It generates a new section, named either "Methods (by class)", "Methods (by generic)" or "Functions".
The section contains a bulleted list describing each function, labelled so that you know what function or method it's talking about.
Here's an example documenting an imaginary new generic:

```{r}
#' Foo bar generic
#'
#' @param x Object to foo.
foobar <- function(x) UseMethod("x")

#' @describeIn foobar Difference between the mean and the median
foobar.numeric <- function(x) abs(mean(x) - median(x))

#' @describeIn foobar First and last values pasted together in a string.
foobar.character <- function(x) paste0(x[1], "-", x[length(x)])
```

### `@rdname`

`@rdname` is a more general-purpose tool.
It overrides the default file name generated by roxygen and merges documentation for multiple objects into one file.
This gives you complete freedom to combine documentation however you see fit.
There are two ways to use `@rdname`.
You can add documentation to an existing function:

```{r}
#' Basic arithmetic
#'
#' @param x,y numeric vectors.
add <- function(x, y) x + y

#' @rdname add
times <- function(x, y) x * y
```

Or, you can create a dummy documentation file by documenting `NULL` and setting an informative `@name`.

```{r}
#' Basic arithmetic
#'
#' @param x,y numeric vectors.
#' @name arith
NULL

#' @rdname arith
add <- function(x, y) x + y

#' @rdname arith
times <- function(x, y) x * y
```

### Order of includes

By default, roxygen blocks are processed in the order in which they appear in the file.
When you're combining multiple files, this can sometimes cause the function usage to appear in a suboptimal order.
You can override the default ordering with `@order`.
For example, the following the block would place `times` first in `arith.Rd` because 1 comes before 2.

```{r}
#' @rdname arith
#' @order 2
add <- function(x, y) x + y

#' @rdname arith
#' @order 1
times <- function(x, y) x * y
```

## Inheriting documentation

You can inherit documentation from other topics in a few ways:

-   `@inheritParams source_function` inherits just the parameter documentation from `source_function()`.

-   `@inherit source_function` will inherit parameters, return, references, description, details, sections, and see also from `source_function`.

-   `@inherit source_function return details` (e.g.) will inherit just return value and details from `source_function()`.

-   `@inheritSection source_function Section title` will inherit the single `@section` called "Section title" from `source_function()`.

All of these tags also work to inherit documentation from functions in another package by using `pkg::source_function`.

## Inline code

To insert code inline, enclose it in `` `r ` ``.
Roxygen will interpret the rest of the text within backticks as R code and evaluate it, and replace the backtick expression with its value.
Here's a simple example:

```{r include = FALSE}
roxygen2:::markdown_on()

simple_inline <- "#' Title `r 1 + 1`
#'
#' Description `r 2 + 2`
foo <- function() NULL
"
```

```{r code=simple_inline}
```

This is equivalent to writing:

```{r code = roxygen2:::markdown(simple_inline)}
```

The resulting text, together with the whole tag is interpreted as markdown, as usual.
This means that you can use R to dynamically write markdown.
For example if you defined this function in your package:

```{r}
alphabet <- function(n) {
  paste0("`", letters[1:n], "`", collapse = ", ")
}
```

You could then write:

```{r echo=FALSE}
env <- new.env()
env$alphabet <- alphabet
roxygen2:::roxy_meta_set("evalenv", env)

backtick <- "#' Title
#' 
#' @param x A string. Must be one of `r alphabet(5)`
foo <- function(x) NULL
"
```

```{r code = backtick}
```

The result is equivalent to writing the following by hand:

```{r code = roxygen2:::markdown_pass1(backtick)}
```

This is a powerful technique for reducing duplication because you can flexibly parameterise the function however best meets your needs.
Note that the evaluation environment is deliberately a child of the package that you're documenting so you can call internal functions.

## Child documents

You can use the same `.Rmd` or `.md` document in the documentation, `README.Rmd`, and vignettes but using child documents:

    ```{r child = "common.Rmd"}`r ''`
    ```

The included Rmd file can have roxygen Markdown-style links to other help topics.
E.g. `[roxygen2::roxygenize()]` will link to the manual page of the `roxygenize` function in roxygen2.
See `vignette("rd-formatting")` for details.

If the Rmd file contains roxygen (Markdown-style) links to other help topics, then some care is needed, as those links will not work in Rmd files by default.
A workaround is to specify external HTML links for them.
These external locations will *not* be used for the manual which instead always links to the help topics in the manual.
Example:

    See also the [roxygen2::roxygenize()] function.

    [roxygen2::roxygenize()]: https://roxygen2.r-lib.org/reference/roxygenize.html

This example will link to the supplied URLs in HTML / Markdown files and it will link to the `roxygenize` help topic in the manual.

Note that if you add external link targets like these, then roxygen will emit a warning about these link references being defined multiple times (once externally, and once to the help topic).
This warning originates in Pandoc, and it is harmless.

## Superseded

Over the years, we have experimented with a number of other ways to reduce duplication across documentation files.
A number of these are now superseded and we recommend changing them to use the techniques described above:

-   Instead of `@includeRmd man/rmd/example.Rmd`, use a child document.

-   Instead of `@eval` or `@evalRd`, use inline R code.

-   Instead of `@template` and `@templateVars` write your own function and call it from inline R code.

Inline R markdown can only generate markdown text within a tag so in principle it is less flexible than `@eval`/`@evalRd`/`@template`.
However, our experience has revealed that generating multiple tags at once tends to be rather inflexible, and you often end up refactoring into smaller pieces so we don't believe this reflects a real loss of functionality.
